Perturbing SU($n$)-symmetric interactions

### 2019-10-30 collective_sun/sun_perturbation.tex
more discussion / derivations in sun_perturbation.tex

%%% symbols, notations, etc.
\usepackage{physics,braket,bm,amssymb} % physics and math
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}[2]{\dfrac{#1}{#2}} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\Braket} % shorthand for braket notation
\renewcommand{\v}{\bm} % bold vectors
\newcommand{\uv}[1]{\bm{\hat{#1}}} % unit vectors
\newcommand{\av}{\vec} % arrow vectors
\renewcommand{\d}{\text{d}} % for infinitesimals
\renewcommand{\c}{\cdot} % inner product

@@

\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}

@@

\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\cov}{cov}
\let\var\relax
\DeclareMathOperator{\var}{var}

@@

As an immediate disclaimer, we note that these notes do {\bf not}
solve the general problem of calculating $H_1,H_2$ for arbitrary
perturbations $\V_1,\V_2$.  Instead, we determine conditions under
which the perturbations $\V_1,\V_2$ generate eigenvectors of the
interaction Hamiltonian $H_0$, and calculate $H_1$ in the case that
these conditions are satisfied.

@@

coupling constants $h_{pq}$ by $\v h$, and a diagonal matrix with
$h_q\equiv\sum_k h_{kq}$ on the diagonals by $\tilde{\v h}$, we thus
find that the vector $\V_1\ket\psi$ is an eigenvector of the
interaction Hamiltonian $H_0$ if the coefficients $\v v$ satisfy the
eigenvalue equation

@@

  \p{\v h - \tilde{\v h}}\c\v v = \Delta\p{\v v} \v v
  \label{eq:v_eig_1}

@@

where $\v x\c\v y$ denotes a standard matrix product of $\v x$ and
$\v y$, and $\Delta\p{\v v}$ is an eigenvalue of
$\v h - \tilde{\v h}$.  If the interaction Hamiltonian $H_0$ is
translationally invariant, then all $h_q=h$, and $\tilde{\v h}=h\,\1$
is proportional to the identity matrix $\1$, so the condition in
\eqref{eq:v_eig_1} becomes

@@

  \v h \c\v v = \sp{\Delta\p{\v v}-h} \v v.

@@

that satisfy \eqref{eq:v_eig_1}, then

@@

We note that vectors $\v v$ satisfying \eqref{eq:v_eig_1} with
$\Delta\p{\v v}\ne 0$ must be mean-zero, as

@@

  v = \f1N \sum_q v_q
  = \f{\sum_{q,k} h_{qk} v_k}{N\sp{\Delta\p{\v v}+h}}
  = \f{\sum_k h_k v_k}{N\sp{\Delta\p{\v v}+h}}
  = \f{h v}{\Delta\p{\v v}+h}
  = \f{v}{\Delta\p{\v v}/h+1},

@@

which is only possible if $\Delta\p{\v v}=0$.  It follows that the
first sum on the right of \eqref{eq:PVVP_start} is

@@

  = - \sum_P v_p^2

@@

To evaluate the operator content of \eqref{eq:PVVP_neq}, we expand

@@

@@

only for
\begin{enumerate*}
\item translationally invariant Hamiltonians, and
\item perturbations $\V_1\p{\v v}$ constructed from vectors $\v v$
  that satisfy the eigenvalue equation in \eqref{eq:v_eig_1}.
\end{enumerate*}

@@

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation in $\V_2$ in
\eqref{eq:perturbations} is
\begin{align}
  H_2^{(1)}
  = \P_0 \V_2 \P_0
  = \sum_{p<q} v_{pq} \P_0 V_p V_q \P_0
  = \f12 \sum_{p\ne q} v_{pq} \P_0 V_p V_q \P_0
  \simeq -\f12 \f{\var\p{\v v}}{N-1} \p{V^2 - W},
\end{align}
where we made use of the result in \eqref{eq:PVVP_neq}.  In order to
compute the second-order effective Hamiltonian $H_2^{(2)}$, as before
we pick an arbitrary fully symmetric state $\ket\psi\in\M_0$ and
expand
\begin{align}
  H \V_2 \ket\psi
  = \sum_{\substack{k<\ell\\p<q}} v_{k\ell} h_{pq}
  \Pi_{pq} V_k V_\ell \ket\psi.
  \label{eq:H_V2_psi_start}
\end{align}
The permutation operator $\Pi_{pq}$ acts trivially on
$V_k V_\ell \ket\psi$ when $p,q\notin\set{k,\ell}$ or
$\set{p,q}=\set{k,\ell}$, leaving a sum of the form
\begin{align}
  \sum_{\substack{p<q\\p,q\notin\set{k,\ell}}} h_{pq} + h_{k\ell}
  = \sum_{p<q} h_{pq}
  - \sum_{\substack{p<q\\p\in\set{k,\ell}\\q\notin\set{k,\ell}}} h_{pq}
  - \sum_{\substack{p<q\\p\notin\set{k,\ell}\\q\in\set{k,\ell}}} h_{pq}
  = E_0 - \sum_{\substack{p\in\set{k,\ell}\\q\notin\set{k,\ell}}} h_{pq}
  = E_0 - 2 h + 2 h_{k\ell},
\end{align}
where we assumed translational invariance with all $h_p=h$.  The
corresponding terms in \eqref{eq:H_V2_psi_start} are then
\begin{align}
  \sum_{k<\ell}
  \sp{\sum_{\substack{p<q\\p,q\notin\set{k,\ell}}} h_{pq} + h_{k\ell}}
  v_{k\ell} V_k V_\ell \ket\psi
  = \p{E_0-2h} \V_2 \ket\psi
  + \sum_{k,\ell} h_{k\ell} v_{k\ell} V_k V_\ell \ket\psi.
\end{align}
The terms in \eqref{eq:H_V2_psi_start} with only one of
$p,q\in\set{k,\ell}$, meanwhile, reduce to
\begin{align}
  \f12 \sum_{\substack{k,\ell\\q\notin\set{k,\ell}}}
  v_{k\ell} \p{h_{kq} V_q V_\ell + h_{\ell q} V_k V_q} \ket\psi
  = \sum_{k,\ell,q} v_{k\ell} h_{kq} V_q V_\ell \ket\psi
  - \sum_{k,\ell} v_{k\ell} h_{k\ell} V_k^2 \ket\psi.
\end{align}
In total,
\begin{align}
  H_0 \V_2 \ket\psi
  = \p{E_0 - 2h} \V_2 \ket\psi
  + \sum_{k,\ell} \p{h_{k\ell} v_{k\ell} + \sum_q h_{kq} v_{q\ell}}
  V_k V_\ell \ket\psi
  - \sum_{k,\ell} h_{k\ell} v_{k\ell} V_k^2 \ket\psi.
  \label{eq:H_V2_psi_mid}
\end{align}
We thus find that the vector $\V_2\ket\psi$ is an eigenvector of the
interaction Hamiltonian $H_0$ if the matrix $\v v$ of all $v_{k\ell}$
satisfies, for some constant $\varepsilon\p{\v v}$,
\begin{align}
  \v h\circ \v v + \v h\c\v v
  = \varepsilon\p{\v v} \v v,
  \label{eq:v_eig_2}
\end{align}
where $\v h\circ\v v$ denotes an element-wise (Hadamard) product of
the matrices $\v h$ and $\v v$, i.e.~with components
$h_{k\ell} v_{k\ell}$; and $\v h\c\v v$ denotes an ordinary matrix
product, with components $\sum_q h_{kq} v_{q\ell}$.  Note that the
diagonal components of the condition in \eqref{eq:v_eig_2} enforce
that the last sum in \eqref{eq:H_V2_psi_mid} vanishes.  If we
construct a perturbation $\V_2\p{\v v}$ with coefficients $\v v$ that
satisfy \eqref{eq:v_eig_2}, then
\begin{align}
  H_0 \V_2\p{\v v} \ket\psi
  = \sp{E_0 - 2h + 2\varepsilon\p{\v v}} \V_2\p{\v v} \ket\psi.
\end{align}

>>>>

%%% symbols, notations, etc.
\usepackage{physics,braket,bm,amssymb} % physics and math
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}[2]{\dfrac{#1}{#2}} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\Braket} % shorthand for braket notation

\renewcommand{\c}{\cdot} % inner product
\renewcommand{\oc}{\circ} % element-wise product

\newcommand{\m}{\bm} % bold symbol
\renewcommand{\v}{\vec} % arrow vector

@@

\renewcommand{\d}{\text{d}} % for infinitesimals
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}

@@

\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\cov}{cov}
\let\var\relax
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\diag}{diag}

@@

As an immediate disclaimer, we note that these notes do {\bf not}
solve the general problem of calculating effective Hamiltonians
$H_1,H_2$ for arbitrary perturbations $\V_1,\V_2$.  Instead, we
\begin{enumerate*}
\item calculate the first-order effective Hamiltonians
  $H_1^{(1)},H_2^{(1)}$,
\item determine conditions under which the perturbations $\V_1,\V_2$
  generate eigenvectors of the interaction Hamiltonian $H_0$, and
\item calculate $H_1$ in the case that these conditions are satisfied
  by $\V_1$.
\end{enumerate*}
When the conditions for $\V_1,\V_2$ to generate eigenvectors of $H_0$
are not satisfied, we hope that the techniques in these notes can be
used to construct eigenvectors and eigenvalues of $H_0$ that will
enable the use of ordinary perturbation theory to determine the
effective Hamiltonians $H_1,H_2$ induced on the fully symmetric
manifold $\M_0$ by the perturbations $\V_1,\V_2$.

@@

coupling constants $h_{pq}$ by $\m h$, and a vector of all
$h_q\equiv\sum_k h_{kq}$ by $\v h$, we thus find that the vector
$\V_1\ket\psi$ is an eigenvector of the interaction Hamiltonian $H_0$
if the coefficients $\v v$ satisfy

@@

  \m h \c \v v - \v h_\d\oc\v v = \Delta\p{\v v} \v v,
  \label{eq:v_cond_1}

@@

where $\m h\c\v v$ is an ordinary matrix-vector product of $\m h$ and
$\v v$, with components $\sum_q h_{kq} v_q$; $\v h_\d\oc\v v$ is an
element-wise (Hadamard) product of $\v h_\d$ and $\v v$, with
components $h_k v_k$; and $\Delta\p{\v v}$ is some constant that may
depend on $\v v$.  Finding vectors $\v v$ that satisfy
\eqref{eq:v_cond_1} is equivalent to finding eigenvectors of the
matrix $\m h-\m h_\d$, where $\m h_\d\equiv\diag\v h_\d$ is a diagonal
matrix; the constant $\Delta\p{\v v}$ is then the eigenvalue of
$\m h-\m h_\d$ corresponding to the eigenvector $\v v$.  If the
interaction Hamiltonian $H_0$ is translationally invariant, then all
$h_q=h$ and $\v h_\d\oc\v v=h\v v$, so the condition in
\eqref{eq:v_cond_1} simplifies to

@@

  \m h \c\v v = \sp{\Delta\p{\v v}+h} \v v.

@@

that satisfy \eqref{eq:v_cond_1}, then

@@

We note that vectors $\v v$ satisfying \eqref{eq:v_cond_1} and
$\Delta\p{\v v}\ne 0$ must be mean-zero, as evident by summing over
the components of \eqref{eq:v_cond_1}:

@@

  \Delta\p{\v v} \sum_k v_k
  = \sum_{k,q} h_{kq} v_q - \sum_k h_k v_k
  = \sum_q h_q v_q - \sum_k h_k v_k
  = 0.
  \label{eq:D_v_0}

@@

It follows that the first sum on the right of \eqref{eq:PVVP_start} is

@@

  = - \sum_p v_p^2

@@

To evaluate the operator content of \eqref{eq:PVVP_neq_start}, we
expand

@@

  \label{eq:PVVP_neq_ops}

@@

only for perturbations $\V_1\p{\v v}$ constructed from vectors $\v v$
that satisfy the eigenvalue equation in \eqref{eq:v_cond_1}.

@@

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation in $\V_2$ in
\eqref{eq:perturbations} is
  H_2^{(1)}
  = \P_0 \V_2 \P_0
  = \sum_{p<q} v_{pq} \P_0 V_p V_q \P_0
  \simeq \f{v}{N-1} \p{V^2 - W},
  v \equiv \f1N \sum_{p<q} v_{pq},
where we made use of the previous result in \eqref{eq:PVVP_neq_ops}.
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand
  H \V_2 \ket\psi
  = \sum_{\substack{k<\ell\\p<q}} v_{k\ell} h_{pq}
  \Pi_{pq} V_k V_\ell \ket\psi
  = \f12 \sum_{\substack{k,\ell\\p<q}} v_{k\ell} h_{pq}
  \Pi_{pq} V_k V_\ell \ket\psi.
  \label{eq:H_V2_psi_start}
The permutation operator $\Pi_{pq}$ acts trivially on
$V_k V_\ell \ket\psi$ when $p,q\notin\set{k,\ell}$ or
$\set{p,q}=\set{k,\ell}$, leaving a sum over $p,q$ of the form
  \sum_{\substack{p<q\\p,q\notin\set{k,\ell}}} h_{pq} + h_{k\ell}
  = \sum_{p<q} h_{pq}
  - \sum_{\substack{p<q\\p\in\set{k,\ell}\\q\notin\set{k,\ell}}} h_{pq}
  - \sum_{\substack{p<q\\p\notin\set{k,\ell}\\q\in\set{k,\ell}}} h_{pq}
  = E_0 - \sum_{\substack{p\in\set{k,\ell}\\q\notin\set{k,\ell}}} h_{pq}
  = E_0 + 2 h_{k\ell} - h_k - h_\ell.
The corresponding terms in \eqref{eq:H_V2_psi_start} with
$p,q\notin\set{k,\ell}$ and $\set{p,q}=\set{k,\ell}$ are then
  E_0 \V_2 \ket\psi + \f12 \sum_{k,\ell}
  \p{2 h_{k\ell} - h_k - h_\ell} v_{k\ell} V_k V_\ell \ket\psi,
while the terms in \eqref{eq:H_V2_psi_start} with only one of
$p,q\in\set{k,\ell}$ simplify to
  \f12 \sum_{\substack{k,\ell\\q\notin\set{k,\ell}}}
  v_{k\ell} \p{h_{kq} V_q V_\ell + h_{\ell q} V_k V_q} \ket\psi
  = \sum_{k,\ell,q} v_{k\ell} h_{kq} V_q V_\ell \ket\psi
  - \sum_{k,\ell} v_{k\ell} h_{k\ell} V_k^2 \ket\psi.
In total,
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_{k,\ell} \p{\sp{h_{k\ell}-\p{\f{h_k+h_\ell}{2}}} v_{k\ell}
    + \sum_q h_{kq} v_{q\ell}} V_k V_\ell \ket\psi
  - \sum_{k,\ell} h_{k\ell} v_{k\ell} V_k^2 \ket\psi.
  \label{eq:H_V2_psi_mid}
We thus find that the vector $\V_2\ket\psi$ is an eigenvector of the
interaction Hamiltonian $H_0$ if the matrix $\m v$ of all $v_{k\ell}$
satisfies, for some constant $\varepsilon\p{\m v}$ that may depend on
$\m v$,
  \tilde{\m h}\oc\m v + \m h\c\m v
  = \varepsilon\p{\m v} \m v,
  \tilde{\m h}
  \equiv \m h - \f12\p{\v h_\d\otimes\v\1 + \v\1\otimes\v h_\d},
  \label{eq:v_cond_2}
where $\tilde{\m h}\oc\m v$ denotes an element-wise (Hadamard) product
of $\tilde{\m h}$ and $\m v$, with components
$\tilde h_{k\ell} v_{k\ell}$; $\m h\c\m v$ denotes an ordinary matrix
product of $\m h$ and $\m v$, with components
$\sum_q h_{kq} v_{q\ell}$; $\v\1$ is a constant vector of ones; and
$\v h_\d\otimes\v\1$ is a tensor product of $\v h_\d$ and $\v\1$.
Note that the diagonal components of the condition in
\eqref{eq:v_cond_2} automatically enforce that the last sum in
\eqref{eq:H_V2_psi_mid} vanishes.  The condition in
\eqref{eq:v_cond_2} can be recast as an eigenvalue equation by
treating the $N\times N$ matrix $\m v$ as a vector with $N^2$
components, which can be further reduced to ${N\choose2}=N\p{N-1}/2$
components by the fact that the matrix $\m v$ is symmetric with zeros
on the diagonal.  If the interaction Hamiltonian $H_0$ is
translationally invariant, then all $h_q=h$ and
$\v h_\d\otimes\v\1\oc\m v=\v\1\otimes\v h_\d\oc\m v=h\m v$, so the
condition in \eqref{eq:v_cond_2} becomes
  \m h\oc\m v + \m h\c\m v = \sp{\varepsilon\p{\m v}+h} \m v,
If we construct a perturbation $\V_2\p{\m v}$ with coefficients $\m v$
that satisfy \eqref{eq:v_cond_2}, then
  H_0 \V_2\p{\m v} \ket\psi
  = \sp{E_0 + 2\varepsilon\p{\m v}} \V_2\p{\m v} \ket\psi.
As a curiosity, we note that if $\V_1\p{\v v}\ket\psi$ is an
eigenvector of $H_0$, then the coefficient vector $\v v$ must satisfy
\eqref{eq:D_v_0}, which we used to determine that $\v v$ was mean-zero
if $\Delta\p{\v v}\ne0$; the analogous condition on matrices $\m v$ in
the two-body case is
  N \varepsilon\p{\m v} v = \f12 \sum_{k,\ell} h_{k\ell} v_{k\ell},
where $v\equiv\sum_{p<q}v_{pq}/N$ is the mean value of the
coefficients in $\V_2\p{\m v}$.
